Columnar VCF for Biobank scale datasets using Zarr

### 2024-04-24 paper.tex
A pass at the intro

\section{Background}
[Four paragraphs. (1) What is VCF, and the good things it has done. 
(2) Explosion in dataset size. Summary of where we're at in terms of biobanks
etc. (3) We have a problem. Existing solutions like TileDB etc
are either proprietary, or tightly coupled to one specific software stack
(Hail, e.g.).
(4) We present the VCF Zarr spec and the vcf2zarr conversion utility.]

@@

>>>>

\section{Background}
Variant Call Format (VCF) is the standard format for interchanging genetic
variation data~\citep{danecek2011variant}, storing information about 
DNA sequence polymorphisms among a set of samples with associated 
quality control metrics and metadata. 
Originally defined specifically as a text file,
it has been refined and standardised~\citep{rehm2021ga4gh} and the 
underlying data-model is now deeply embedded in bioinformatics practice.
Dataset sizes have grown explosively since the introduction of 
VCF as part of 1000 Genomes project~\citep{10002015global},
with Biobank-scale projects such as 
Genomics England~\cite{turnbull2018100},
UK Biobank~\citep{bycroft2018genome,backman2021exome,halldorsson2022sequences,uk2023whole},
and the All of Us research program~\citep{all2024genomic}
collecting genome sequence data for hundreds of thousands of humans.
While there are many advantages to encoding data in text files,
VCF is simply not fit for purpose as part of computational workflows 
at this scale. [A sentence somehow establishing that methods analysing 
large-scale genomic data are held back by VCF, needing to first 
convert to intermediate formats like tsinfer.]

The main solution that has emerged to these problems of 
computation on genetic variation data at scale is Hail~\cite{ganna2016ultra,hail2024}.
Building on components from the Hadoop ecosytem~\citep{white2012hadoop},
Hail distributes computations transparently over a cluster, enabling
interactive analysis of very large datasets. It has been successfully used 
in many studies, and instrumental in generating datasets such as 
gnomadAD~\cite{karczewski2020mutational,chen2024genomic}.
While Hail is powerful software, it is not a suitable platform for 
developing inference or performing complex analysis. It is complex,
vertically integrated software, with an undocumented file format
[say this better]. 
Similarly, commerical solutions that have emerged to facilitate
the analysis of large-scale genetic variation data are either
based on proprietary~[Terra? BigQuery?]
or single-vendor technologies~\cite[e.g.][]{tiledb2024}.

VCF has enabled a vibrant ecosystem of interoperating
utilities~\cite{danecek2021twelve,garrison2022spectrum}, but these tools
are fundamentally scale-limited by the format. 
The most intrinsically limiting aspect of VCF's design is its 
row-wise layout of data within a file, which means that (for example)
data for a particular sample cannot be efficiently retrieved.
The file-oriented paradigm is also unsuited to the realities 
of modern datasets, which are too large to download and 
often required to stay in-situ by data-access agreements.
Large files are stored in cloud environments, where the 
file systems that are required by file-oriented tools
are expensively emulated on the basic building blocks
of object storage. 
These multiple layers of inefficiences around processing
VCF data at scale in the cloud mean that it is 
time-consuming and expensive, effectively making 
% Dunno if we want to get into this, but it's true
analysis of these datasets the preserve of a well-funded elite.

Here, we decouple the VCF data model from its row-oriented
file definition, and show how the corresponding data model can be 
efficiently stored and analysed in a cloud-native manner.
We do this by translating VCF data into Zarr format,
a popular method of storing compressed N-dimensional arrays
used to store petabyte-scale datasets across the sciences [citations].
We present the VCF Zarr specification which formalises this 
mapping, and and the \texttt{vcf2zarr} 
utility which reliably converts large-scale VCFs into Zarr 
format. We show that VCF Zarr is much more compact than 
VCF (and BCF), and it competitive with state-of-the-art
VCF compression tools. Moreover, we show that Zarr's 
storage of data in an analysis-ready 
format~\cite{abernathey2021cloud} greatly facilitates computation,
with various benchmarks being substantially faster than
\texttt{bcftools} based pipelines, and again competetive
with state-of-the-art methods. Finally, we show the 
utility of VCF Zarr on the Genomics England aggV2 dataset,
demonstrating that common queries can be performed orders
of magnitude more quickly based on simple Python scripts.

@@

Although Zarr is optimised for cloud settings, it also
works very well in the classical (shared) file-system setting
and we therefore focus on benchmarks in this setting
to allow us to compare with existing tools.
Those seeking benchmarks of Zarr in cloud deployments are 
directed to [citations].
